Annotated example of a spatially varying effect with sdmTMB

### 2021-11-26 doc/appendix-snow-sdmTMB.Rmd
clean up owl appendix

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
theme_set(theme_minimal())
```

This illustrates the fitting of a model with a spatially varying coefficient for the effect a regional climate index on species count data using sdmTMB. 
We will use counts of Snow Owls (*Bubo scandiacus*) observed in North America during annual Christmas Bird Counts (CBCs) between 1979 and 2020. 
This citizen science survey is organized by the by the National Audubon Society (data available by request from cbcadmin@audubon.org). 
Each survey is conducted between December 14 to January 5 in 24.14 kilometre diameter circle. 
Circle locations are chosen by local volunteers and therefore reflect both locations of higher human densities and known bird watching locations.
Each circle (given a unique "CBCID") is generally surveyed annually by at least 10 volunteers, or deemed inactive.

@@

```{r load-data, cache=FALSE} 
dat <- readRDS(here::here("owls/SNOW_data.rds"))
glimpse(dat)
```
Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km.

@@

```{r mesh, echo=TRUE, cache=FALSE}
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 1.5)
```

```{r plot-mesh, fig.asp=0.9, out.width="4in"}
plot(mesh)
```
The result is a mesh with `r mesh$mesh$n` knots.

Next, we will plot the counts through time to see if there is an obvious overall trend, or relationship with NAO.

```{r explore, fig.width = 8}
ggplot(dat, aes(year, log(count+1), colour = nao)) + 
  geom_jitter(alpha = 0.5) + 
  scale_colour_viridis_c()
```

Because there does not appear to be much correlation between years, year will be included as a random factor: `(1|year_f)`.

We will also include a fixed spatial random field (`spatial = "on"`) to account for consistent differences in habitat, proximity to the breeding grounds, and effects of flyways and barriers like coasts, and independent and identically distributed spatiotemporal random field (`spatiotemporal = "IID"`) to allow for annual deviations in these spatial effects due to factors like local weather and prey abundance. 
The random intercept for year allows the mean estimate for each of these spatiotemporal fields to vary. 
If years in sequence were strongly correlated with each other, we could use a random walk to link their mean estimates by setting `time_varying = ~ 1`.

Our only fixed effect will be the NAO climate index (`nao`), but because of the wide spatial extent of these surveys, we also expect that the effect of NAO might not be consistent throughout North America. 
Therefore, effect of NAO will also be allowed to vary spatially using this formula: `spatial_varying = ~ 0 + nao`.

First, we will try fitting GLMMs (generalized linear mixed effects models) with a few different distributions appropriate for count data. 
We will use `normalize = TRUE` to speed up optimization. 
Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output.

```{r fit-poisson, echo=TRUE, eval=FALSE}
m0 <- sdmTMB(count ~ 1 + nao + (1|year_f),
             time = "year",
             spatial_varying = ~ 0 + nao,
             spatial = "on", 
             spatiotemporal = "IID",
             family = poisson(link = "log"),
             control = sdmTMBcontrol(normalize = TRUE),
             data = dat, mesh = mesh)
```

```{r fit-negb1, echo=TRUE, eval=FALSE}
m1 <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom1(link = "log"),
            control = sdmTMBcontrol(normalize = TRUE),
            data = dat, mesh = mesh)
```

```{r fit-negb2, echo=TRUE, eval=FALSE}
m2 <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom2(link = "log"),
            control = sdmTMBcontrol(normalize = TRUE),
            data = dat, mesh = mesh)
```

```{r load-models, warning=FALSE, message=FALSE, echo=FALSE}
# m0 <- readRDS(here::here("owls/snow_w_main_effect_0_150km.rds"))
# m1 <- readRDS(here::here("owls/snow_w_main_effect_1_150km.rds"))
# m2 <- readRDS(here::here("owls/snow_w_main_effect_150km.rds"))
m <- readRDS(here::here("owls/snow_w_main_effect_150km_reml.rds"))
```

```{r aic, echo=TRUE, cache=TRUE}
# AIC(m0,m1,m2)
```

The `nbinom2` distribution appears to have the best fit. 
We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and then check our residuals. 
This method of checking residuals is not guarenteed to be accurate, so an additional step using STAN could be added.

```{r fit-negb2-reml, echo=TRUE, eval=FALSE}
m <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom2(link = "log"),
            reml = TRUE,
            data = dat, mesh = mesh)
```

We can predict on the original data locations:

```{r predict}
pred <- predict(m)
```

We can inspect randomized quantile residuals:

```{r qqnorm, out.width="5in"}
pred$resid <- residuals(m)
qqnorm(pred$resid)
qqline(pred$resid)
```

```{r}
library(tmbstan)
s <- tmbstan(m$tmb_obj, iter = 200, chains = 1, warmup = 199)
pstan <- predict(m, tmbstan_model = s)
resid <- as.numeric(pstan) - m$data$observed
# or do any PIT residuals randomization here...
qqnorm(resid);qqline(resid) # should be right
```

```{r plot-resids}
ggplot(pred, aes(X, Y, colour = resid)) +
  geom_point(size = 0.5) +
  facet_wrap(~year) +
  coord_fixed() +
  scale_colour_gradient2()
```

@@

```{r print-fit}
print(m)
```

Extract the parameter estimates as a data frame:

@@

In order to produce pretty maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and reproject. 
One can also get continental coastlines from the package rnaturalearth and outlines of the great lakes from   http://www.naturalearthdata.com/downloads/10m-physical-vectors/ and transform them to the correct projection using `sf::st_transform(crs = proj)`.

```{r proj-p, echo=TRUE, eval=FALSE}
p <- pred %>% mutate(
  X = X * 100000,
  Y = Y * 100000
)

@@

@@

```{r p-mean, echo=TRUE, eval=FALSE}
b <- tidy(m)

@@

```{r plot-map, echo=TRUE, eval=FALSE}
ggplot(data = p_proj) +
  geom_point(data = p_mean, aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), # adjust buffers slightly
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "top", label.position = "bottom")) +
  guides(size = "none") + labs(colour = "NAO effect\non Snowy Owl count") +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

There are four other kinds of predictions that we get out of the model. 
We can plot our predictions of winter owl counts that incorporate all fixed effects and random effects:

```{r plot-all-effects, echo=TRUE, eval=FALSE}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(
    # transform to make variation in low number visible 
    trans="sqrt", 
    # and also trim the top extremes of the colour axis 
    limits = c(0, quantile(exp(p_proj$est), 0.995)), na.value = "yellow") +
  labs(x = "Longitude", y = "Latitude", colour = "Predicted\ncount") +
  # theme_bw() + 
  theme(axis.text = element_blank(), axis.ticks = element_blank(), axis.title = element_blank()) +
  facet_wrap(~year_f)
```

Note that for these annually varying estimates, we are showing only the most recent 25 years. 
We can also look at just the fixed effects, but in this case these only include the overall effect of mean annual NAO:

```{r plot-fixed-effects}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est_non_rf)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c() + 
  # theme_bw() + 
  theme(axis.text = element_blank(), axis.ticks = element_blank(), axis.title = element_blank()) +
  facet_wrap(~year_f)
```

We can look at the spatial random effects that represent consistent deviations in space that are not accounted for by our fixed effects. 
In other words, these deviations represent consistent biotic and abiotic factors that are affecting winter owl counts.
These could be landcover, proximity to the breeding grounds, and effects of flyways and barriers like coasts.

```{r plot-spatial-effects, echo=TRUE, eval=FALSE}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), # adjust buffers slightly
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "top", label.position = "bottom")) +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

Finally, we can look at the spatiotemporal random effects that represent deviations from the fixed effect predictions and the spatial random effect deviations. 
These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model. 
For owls, this is most likely to be things like landcover changes, prey abundance cycles, and snow depth.

```{r plot-spatiotemporal-effects}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = epsilon_st), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_fill_gradient2() +
  # theme_bw() + 
  theme(axis.text = element_blank(), axis.ticks = element_blank(), axis.title = element_blank()) +
  facet_wrap(~year_f)
```

>>>>

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
# library(sf)
theme_set(theme_bw())
# theme_set(theme_minimal())
```

This illustrates the fitting of a model with a spatially varying coefficient for the effect a regional climate index on species count data using sdmTMB. 
We will use counts of Snow Owls (*Bubo scandiacus*) observed in North America during annual Christmas Bird Counts (CBCs) between 1979 and 2020. 
This citizen science survey is organized by the by the National Audubon Society (data available by request from cbcadmin@audubon.org). 
Each survey is conducted between December 14 to January 5 in 24.14 kilometre diameter circle. 
Circle locations are chosen by local volunteers and therefore reflect both locations of higher human densities and known bird watching locations.
Each circle (given a unique "CBCID") is generally surveyed annually by at least 10 volunteers, or deemed inactive. 
While these locations are not random, the consistent sampling through time still make them useful for exploring spatiotemporal processes.

@@

```{r load-data, cache=FALSE} 
snow <- readRDS(here::here("owls/SNOW_data.rds"))
glimpse(snow)
```

Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km.

@@

```{r mesh, echo=TRUE, cache=FALSE}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
```

```{r plot-mesh, fig.asp=0.9, out.width="6in"}
plot(mesh)
```

The result is a mesh with `r mesh$mesh$n` knots.

Next, we will plot the owl counts through time to see if there is an obvious overall trend, or relationship with NAO.

```{r explore, fig.asp=0.4, fig.width = 6}
ggplot(snow, aes(year, count, colour = nao)) + 
  geom_jitter(alpha = 0.5) + 
  scale_colour_viridis_c()
```

# Model fitting

First, we will try fitting GLMMs (generalized linear mixed effects models) with a few different distributions appropriate for count data. 
<!-- We will use `normalize = TRUE` to speed up optimization.  -->
Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output.

Our only fixed effect will be the NAO climate index (`nao`), but because of the wide spatial extent of these surveys, we also expect that the effect of NAO might not be consistent throughout North America. 
Therefore, effect of NAO will also be allowed to vary spatially using this formula: `spatial_varying = ~ 0 + nao`.
Because there does not appear to be much correlation between years, year will be included as a random factor: `(1|year_f)`. 
We will also include a fixed spatial random field (`spatial = "on"`) to account for consistent differences in habitat, proximity to the breeding grounds, and effects of flyways and barriers like coasts, and independent and identically distributed spatiotemporal random field (`spatiotemporal = "IID"`) to allow for annual deviations in these spatial effects due to factors like local weather and prey abundance. 
The random intercept for year allows the mean estimate for each of these spatiotemporal fields to vary. 
If years in sequence were strongly correlated with each other, we could use a random walk to link their mean estimates by setting `time_varying = ~ 1`.

```{r fit-poisson, echo=TRUE, eval=FALSE}
m0 <- sdmTMB(count ~ 1 + nao + (1|year_f),
             time = "year",
             spatial_varying = ~ 0 + nao,
             spatial = "on", 
             spatiotemporal = "IID",
             family = poisson(link = "log"),
             data = snow, mesh = mesh)
```

```{r fit-negb1, echo=TRUE, eval=FALSE}
m1 <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom1(link = "log"),
            data = snow, mesh = mesh)
```

```{r fit-negb2, echo=TRUE, eval=FALSE}
m2 <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom2(link = "log"),
            data = snow, mesh = mesh)
```

```{r load-models, warning=FALSE, message=FALSE, echo=FALSE}
m0 <- readRDS(here::here("owls/snow_w_main_effect_0_150km.rds"))
m1 <- readRDS(here::here("owls/snow_w_main_effect_1_150km.rds"))
m2 <- readRDS(here::here("owls/snow_w_main_effect_150km.rds"))
m <- readRDS(here::here("owls/snow_w_main_effect_150km_reml.rds"))
```

```{r aic, echo=TRUE, cache=TRUE}
AIC(m0,m1,m2)
```

The `nbinom2` distribution appears to have the best fit. 
We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and then check our residuals.

```{r fit-negb2-reml, echo=TRUE, eval=FALSE}
m <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom2(link = "log"),
            reml = TRUE,
            data = snow, mesh = mesh)
```

@@

```{r print-fit, message=FALSE}
print(m)
```

View the confidence intervals on both fixed and random parameter estimates using `tidy()`.

@@

We can also renerate predictions and residuals for the original data locations:

```{r predict, message=FALSE}
pred <- predict(m)
```

We can inspect randomized quantile residuals, although this method of checking residuals is not guaranteed to be accurate. 
An additional step using STAN could be added.

```{r qqnorm, fig.asp=0.9, out.width="3in"}
pred$resid <- residuals(m)
qqnorm(pred$resid)
qqline(pred$resid)
```

```{r plot-resids, fig.width = 7}
ggplot(pred, aes(X, Y, colour = resid)) +
  geom_point(size = 0.5) +
  facet_wrap(~year) +
  coord_fixed() +
  scale_colour_gradient2()
```

# Plotting spatial predictions

In order to produce pretty maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will appear in latitude and longitude.

```{r proj-p, echo=TRUE, eval=T}
library(sf)

p <- pred %>% mutate(
  X = X * 100000,
  Y = Y * 100000
)

@@

One can also get continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http://www.naturalearthdata.com/) and transform them to the correct projection using `sf::st_transform(crs = proj)`.

```{r shapes, echo=FALSE, message=FALSE}
library("rnaturalearth")
library("rnaturalearthdata")
coast <- ne_coastline(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(crs = proj)
lakes <- sf::st_read(here::here("owls/ne_10m_lakes"))
lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = proj) # only large lakes
```

## Spatially varying effect

@@

```{r p-mean, message=FALSE, echo=TRUE, eval=T}
b <- tidy(m)

@@

```{r plot-map, echo=TRUE, eval=T}
ggplot(data = p_proj) +
  geom_point(data = p_mean, 
     aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top", label.position = "bottom")) +
  guides(size = "none") + 
  labs(colour = "NAO effect\non Snowy Owl count") +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

## Predicted counts

These are the predictions of winter owl counts that incorporate all fixed effects and random effects:

```{r plot-all-effects, echo=TRUE, eval=T, fig.width = 7}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(
    # transform to make variation in low number visible 
    trans="sqrt", 
    # and also trim the top extremes of the colour axis 
    limits = c(0, quantile(exp(p_proj$est), 0.995)), na.value = "yellow") +
  labs(x = "Longitude", y = "Latitude", colour = "Predicted\ncount") +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        axis.title = element_blank()) +
  facet_wrap(~year_f)
```
Note that for these annually varying estimates, we are showing only the most recent 25 years.

## Other components of predicted counts

In addition to the spatially varying effect we started with, there are three other components that go into the overall predictions that we can plot spatially.

1. Just fixed effects

<!-- confirm that the random effect of year is not included here? -->
In this case these only include the overall effect of mean annual NAO:

```{r plot-fixed-effects, echo=TRUE, eval=T}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est_non_rf)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c() + 
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        axis.title = element_blank()) +
  facet_wrap(~year_f)
```

2. Static spatial random effects

We can look at the spatial random effects (`omega_s`) that represent consistent deviations in space that are not accounted for by our fixed effects. 
In other words, these deviations represent consistent biotic and abiotic factors that are affecting winter owl counts.
These could be landcover, proximity to the breeding grounds, and effects of flyways and barriers like coasts.

```{r plot-spatial-effects, echo=TRUE, eval=T}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top", label.position = "bottom")) +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

3. Spatiotemporal random effects

Finally, we can look at the spatiotemporal random effects (`epsilon_st`) that represent deviations from the fixed effect predictions and the spatial random effect deviations. 
These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model. 
For owls, this is most likely to be things like landcover changes, prey abundance cycles, and snow depth.

```{r plot-spatiotemporal-effects, echo=TRUE, eval=T}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = epsilon_st), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c() +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        axis.title = element_blank()) +
  facet_wrap(~year_f)
```
